\documentclass[11pt]{report}

% Document dimensions
\usepackage{geometry}
\geometry{top=1.5cm, bottom=1.5cm, textwidth=15cm}

% Math related packges.
\usepackage{amsmath}
\usepackage{cancel}

% Natural Deduction package
\usepackage{proof}
\usepackage{mdframed}

% Fix the header space: start at the top of the page.
\usepackage{hyperref}

% Import the necessary preamble for the document. 
\usepackage{../../proofsPrograms}


\begin{document}
	
	
% Heading for the tutorial	
\begin{center}
	{\bf MATH230: Tutorial Six}
\end{center}
\begin{center}
	{\bf Computation and Church Encodings}
\end{center}


% Box with goals and relevent lecture notes.
\noindent\fbox{
	\parbox{\textwidth}{

		Key ideas
			\begin{itemize}
				\item Practice $\beta$-reduction, 
				\item Encode logic in $\lambda$-calculus, 
				\item Encode natural numbers in $\lambda$-calculus.
				\item Encode compound data in $\lambda$-calculus.
			\end{itemize}

		Relevant topic: Untyped Lambda Calculus Slides\\
		Relevant reading: Type Theory and Functional Programming, Simon Thompson
		
	\vspace{0.2cm}

	Hand in exercises: 1b, 2, 3b, 4c, 5b\\ 
	{\bf Due Friday @ 5pm to the submission box on Learn.}
	}
}
% Discussion questions for tutor.
\newline
\vspace{0.5cm}

\noindent {\bf Discussion Questions}

\begin{itemize}


	\item Compute the normal forms of the following $\lambda$-terms: 
	
	\begin{itemize}
		\item[] $(\lambda x. \ x \ x)(\lambda y. \ \lambda z. \ y \ z \ z)$
		\vspace{70mm}
		\item[] AND TRUE TRUE
	
	\end{itemize}
	
\end{itemize}	

% New page for tutorial exercises.
\newpage
{\bf Tutorial Exercises}
\begin{enumerate}
	
	\item Compute the normal form for each of the following $\lambda$-terms:
	
		\begin{enumerate}
			\item NOT FALSE
			\item OR TRUE FALSE
			\item AND FALSE TRUE 
			\item IMPLIES FALSE TRUE 
		\end{enumerate} 

	{\bf Only expand those expressions necessary for each step.}
		
	\item Write down $\lambda$-expressions that represent the propositional binary connectives XOR, NAND, and NOR. Recall that these have the following truth tables.
			
	\vspace{0.5cm}

	\begin{center}
		$\begin{array}{c c c}

			\begin{array}{ c c | c }			
				P & Q & \text{XOR}(P,Q)\\
				\cline{1 - 3}
				T & T & F \\ 
				T & F & T \\ 
				F & T & T \\ 
				F & F & F
			\end{array} 
			& \hspace{0.5cm}
			\begin{array}{ c c | c }			
				P & Q & \text{NAND}(P,Q)\\
				\cline{1 - 3}
				T & T & F \\ 
				T & F & T \\ 
				F & T & T \\ 
				F & F & T
			\end{array} 
			& \hspace{0.5cm}
			\begin{array}{ c c | c }			
				P & Q & \text{NOR}(P,Q)\\
				\cline{1 - 3}
				T & T & F \\ 
				T & F & F \\ 
				F & T & F \\ 
				F & F & T
			\end{array} 
		\end{array}$
	\end{center}
	\vspace{0.2cm}

	\item By substituting the explicit $\lambda$-expressions (as necessary) and performing $\beta$-reduction, determine the normal forms of the following $\lambda$-expressions. 

	\begin{enumerate}
		\item SUCC ONE
		\item SUM ONE TWO
		\item MULT TWO THREE
	\end{enumerate}

	{\bf Only expand those expressions necessary for each step.}

	\item We have defined the following $\lambda$-expression to construct pairs of $\lambda$-expressions:
	
	$$\text{PAIR} = \lambda x. \ \lambda y. \ \lambda f. \ f \ x \ y$$

	The third input is a built-in place ready to take a selector:

	$$\text{FirST} = \lambda x. \ \lambda y. \ x \hspace{1cm} \text{SecoND} = \lambda x. \ \lambda y. \ y$$

	 Reduce these to normal form 

		\begin{enumerate}
			\item PAIR a b FST \vspace{0.2cm}
			\item PAIR a b SND \vspace{0.2cm}
			\item PAIR (PAIR a b) (PAIR c d) SND
		\end{enumerate}

%	\item Integers are solutions to equations $x + b = a$. We can represent natural number solutions using Church numerals. However, further abstractions are required to represent the negative solutions to such equations. 
%
%	{\bf Example} The equation $x + 1 = 0$ has solution $x = -1$. One way to represent this in the $\lambda$-calculus is to use the pair $(0,1)$ which we interpret as $-1 = (0,1)$. 
%	
%	More generally $k = (a,b) = a - b$ represents the solution to $x + b = a$.
%
%	\hspace{0.5cm} {\bf Such representations are not unique!} e.g. $0 = (0,0) = (1,1) = \dots$
%
%	Write $\lambda$-expressions for arithmetic on integers as pairs of Church numerals.
%		\begin{itemize}
%			\item[] INT-SUM to calculate the sum of two integers.
%			\item[] INT-MULT to calculate the product of two integers.
%			\item[] INT-NEG(ative) to calculate the negative of an integer.
%		\end{itemize}

	\item Positive rational numbers are solutions to equations of the form $bx = a$, where $a,b : \mathbb{N}$. Use PAIR to represent positive rational numbers in the $\lambda$-calculus and write $\lambda$-expressions to compute rational number arithmetic. 
		\begin{itemize}
			\item[(a)] RAT-SUM to calculate the sum of two rational numbers.
			\item[(b)] RAT-MULT to calculate the product of two rational numbers.
			\item[(c)] RAT-REC(iprocal) to calculate the reciprocal of an integer.
		\end{itemize}

%	\item Imaginary numbers can be represented as PAIRs of rational numbers.
%	
%	\item Real numbers are a more complicated issue.

\end{enumerate}
	
\end{document}